Appendix: Trends in age-adjusted mortality rates for blacks and whites by cause and sex, 1969-2013, United States

### 2017-03-30 Manuscript/Manuscript_COD_Trends.Rmd

**The data**
We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia between 1969 and 2013. The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System. The population data were derived by the NCHS using intercensal bridged-race population estimates. This study used data on the number of deaths and the population size according to race (black or white), sex (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 -- 2013), and cause of death. Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013). These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes (**S1 Table**). When the number of deaths for a population stratum is between 1 and 9, this data point is suppressed by the software and produces a missing death count for that cell However, death counts of exactly 0 or greater than 9 are known precisely.

@@

**Statistical Methods**
The data structure produced two statistical challenges that were addressed by our quantitative approach. First, our methodology needed to impute the suppressed death counts (for counts between 1 and 9). Second, several states had small black populations which produced sparse counts in each sex-age category. In these cases, the estimation of the sex-age mortality rates is imprecise, and can produce abrupt changes in the estimated mortality rates over time that are unrealistic. To better estimate the underlying age-specific mortality rates, a method that produces smoother age-specific mortality trends was required.

To address these challenges, we used a Bayesian autoregressive model to smooth the data over time, and incorporated a truncated Poisson model to impute the suppressed death counts. Specifically, we used Poisson regression to model the number of deaths in each state-age-sex-race-year category as a function of the at-risk population. The mortality rates over time within each state-age-sex-race category are modeled autoregressively, where the log mortality rate in the current year was modeled as a normal variable centered at the log mortality rate estimated from the previous year. Truncated Poisson regression was used to impute the suppressed death counts, by specifying that the suppressed counts were sampled from a Poisson distribution and were constrained to be between 1 and the maximum of 9 and the at-risk population size (which may be smaller than 9).

@@

To calculate life expectancy at birth, the smoothed death counts were aggregated to estimate the total deaths in each age group and standard life table methods were applied [ref]. Life expectancy at birth (in years) was calculated for blacks and whites in each state and year according to sex. The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in each state, year, and sex.

@@

To describe the general pattern in these trend lines, we regressed each state's mean contribution against a linear spline for calendar time that was estimated with five degrees of freedom. This model also contained fixed effects for state and was inversely-weighted by the standard error of each state's mean estimate within each year. We then calculated residuals for each year by substracting the meta-estimated contribution from the smoothed-observed estimated contribution and summed these residuals over time for each state to rank states according to how well the model predicted their pattern. Plots of these residuals over time were used to identify divergent states.

**Need to add information on the trends in mortality**

###Results

>>>>

**Description of the mortality data**
We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia between 1969 and 2013. The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System. The population data were derived by the NCHS using intercensal bridged-race population estimates. This study used data on the number of deaths and the population size according to race (black or white), sex (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 -- 2013), and cause of death. Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013). These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes (**S1 Table**). When the number of deaths for a population stratum is between 1 and 9, this data point is suppressed by the software and produces a missing death count for that cell However, death counts of exactly 0 or greater than 9 are known precisely.

@@

**Statistical challenges to be addressed**
The data structure produced two statistical challenges that were addressed by our quantitative approach. First, our methodology needed to impute the suppressed death counts (for counts between 1 and 9). Second, several states had small black populations which produced sparse counts in each sex-age category. In these cases, the estimation of the sex-age mortality rates is imprecise, and can produce abrupt changes in the estimated mortality rates over time that are unrealistic. To better estimate the underlying age-specific mortality rates, a method that produces smoother age-specific mortality trends was required.

**Estimation of smoothed mortality rates using a Bayesian autoregressive model**
To address these challenges, we used a Bayesian autoregressive model to smooth the data over time, and incorporated a truncated Poisson model to impute the suppressed death counts. Specifically, we used Poisson regression to model the number of deaths in each state-age-sex-race-year category as a function of the at-risk population. The mortality rates over time within each state-age-sex-race category are modeled autoregressively, where the log mortality rate in the current year was modeled as a normal variable centered at the log mortality rate estimated from the previous year. Truncated Poisson regression was used to impute the suppressed death counts, by specifying that the suppressed counts were sampled from a Poisson distribution and were constrained to be between 1 and the maximum of 9 and the at-risk population size (which may be smaller than 9).

@@

**Decomposition of the life expectancy gap by cause of death**
To calculate life expectancy at birth, the smoothed death counts were aggregated to estimate the total deaths in each age group and standard life table methods were applied [ref]. Life expectancy at birth (in years) was calculated for blacks and whites in each state and year according to sex. The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in each state, year, and sex.

@@

**Description of the trend in each cause of death's contribution to the gap and identification of divergent states**
To describe the general pattern in these trend lines, we regressed each state's mean contribution against a linear spline for calendar time that was estimated with five degrees of freedom. This model also contained fixed effects for state and was inversely-weighted by the standard error of each state's mean estimate within each year. We then calculated residuals for each year by substracting the meta-estimated contribution from the smoothed-observed estimated contribution and summed these residuals over time for each state to rank states according to how well the model predicted their pattern. Plots of these residuals over time were used to identify divergent states.

*Need to add information on the trends in mortality* -- where?

###Results
```{r initialize, echo=FALSE, warning=F, message=F}
library(shiny)
library(plotly)
library(viridis)
library(scales)
library(shinythemes)
library(dplyr)
library(purrr)
library(stringr)
library(DT)
source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
```

```{r load_data, echo=F}
BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_results.csv")
BlackWhite_results <- reorder.as.map(BlackWhite_results, "state")
```

```{r life-expectancy-trends, echo=F}
LE_trend_plot_males <- ggplot(data = subset(BlackWhite_results, sex == "Male"), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
  facet_wrap(~ state.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Life expectancy gap (years)") +
  xlab("Year (1969-2013)")

LE_trend_plot_females <- ggplot(data = subset(BlackWhite_results, sex == "Female"), aes(y = LE_wbgap_mean, x = year)) + 
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
  facet_wrap(~ state.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Life expectancy gap (years)") +
  xlab("Year (1969-2013)")
```

```{r plot_both, eval=F}
#Informative: male and females at the same time
ggplotly(ggplot(data = subset(BlackWhite_results), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, group = sex, fill = Census_Division), alpha = 0.5) +
  geom_line(aes(group = sex, col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
  facet_wrap(~ state.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Life expectancy gap (years)") +
  xlab("Year (1969-2013)"))
```

**Figures 1 and 2** display trends in the life expectancy gap for males and females, respectively. For both sexes, the life expectancy gap has decreased since 1969, although several states exhibited a period of increasing inequality that peaked in the early 1990s, especially among males.

**Figures 2 - 14** display trends in the contribution of each cause of death to the life expectancy gap for males and females, in years. The green band illustrates the credible interval for the smoothed contribution estimated for each state, and illustrates the heterogeneity across states in their contributions trends. These curves are overlaid with the meta-estimated common trend, shown with a red line, and the plots are organized according to the closeness of the [fit] (*need another word*) of the red curve to the green band. Thus, increasingly divergent states are shown further down the display. We supplement these main results with smoothed estimates of the age-standardized mortality rates for each cause of death (**eFigures 1-12**).

**Cardiovascular disease**

Cardiovascular disease is the most common cause of death for both men and women of both races. Mortality from cardiovascular disease decreased markedly since 1969 for both sexes and races (**eFigures 1 and 2**). However, white males underwent relatively larger mortality decreases than black males, resulting in a modest increase to the contribution of cardiovascular disease to the life expectancy gap for most states (**Figure 2**). In particular, Pennsylvania, Michigan, Wisconsin, and Washington DC experienced relatively larger increases to their CVD contributions to the gap for males, while some states such as Illinois and Kentucky exhibited relatively larger early contributions in the 1970s.

The situation was reversed for females, with black females experiencing relatively larger mortality decreases over time compared to white females. Thus, the contribution of CVD to the life expectancy gap decreased by more than a year in several states. Of note, some states experienced a period of early increasing contributions, most notably, Arizona, Washington, and Oregon, that was not captured by the general pattern. Furthermore, Wisconsin and Washington DC both exhibit increasing contributions over time, as mortality gains among white females outpaced those among black females.

**Cancer**

**Figure 1: State-level trends in the difference in life expectancy between blacks and whites in males, United States, 1969-2013**

```{r plot_LE_trends_males, echo=F, fig.height=9, fig.width=11}
ggplotly(LE_trend_plot_males)
```

**Figure 2: State-level trends in the difference in life expectancy between blacks and whites in females, United States, 1969-2013**

```{r plot_LE_trends_females, echo=F, fig.height=9, fig.width=11}
ggplotly(LE_trend_plot_females)
```

###Discussion

###Appendix

```{r, echo = F}
#cant figure out how to do this yet (embed other Rmd file inside this one)
knitr::knit('/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Manuscript/Appendix-agestd-mort-rates.Rmd', tangle = T)
source('/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Manuscript/Appendix-agestd-mort-rates.R')
```
